---
title: "Working with MSstatsConvert"
output: rmarkdown::html_vignette
vignette: >
 %\VignetteIndexEntry{Working with MSstatsConvert}
 %\VignetteEngine{knitr::rmarkdown}
 %\VignetteEncoding{UTF-8}
---
  
```{r, include = FALSE}
  knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
  )
```
  
  
# Purpose of MSstatsConvert
  
The `MSstatsConvert` package is a member of the `MSstatst` family of packages, MSstats and MSstatsTMT.
It creates an abstraction for the steps in mass spectrometry (MS) data analysis 
that are required before a dataset can be used for statistical modeling. 
In short, the package is responsible for converting output from signal processing
tools such as `OpenMS` or `MaxQuant` into a format suitable for statistical analysis. 
This includes:
  
  - PSM- and protein-level filtering,
  - managing shared peptides,
  - removing decoy, iRT, and other irrelavant sequences,
  - removing features or proteins with a low number of measurements,
  - aggregating duplicated measurements,
  - handling fractionation by removing overlapped features,
  - creating balanced statistical design in the presence of missing data.
  
`MSstatsConvert` allows for transforming any MS quantification result into a format
required by `MSstats` and `MSstatsTMT` packages.
Additionally, it provides built-in cleaning functions for outputs of `DIAUmpire`, `MaxQuant`,
`OpenMS`, `OpenSWATH`, `Progenesis`, `ProteomeDiscoverer`, `Skyline`, `Spectromine`, and `Spectronaut`.
These functions serve as a base for converter functions (called `*toMSstatsFormat` or 
`*toMSstatsTMTFormat`) provided  by the `MSstats` and `MSstatsTMT` packages.
  
# MSstats data format
  
MSstats family packages works with label-free, SRM and TMT datasets. 
The following column are required.

  - `ProteinName`: column that indicates a protein ID. If the analysis is to be made 
  at the peptide-level, the column should store peptide IDs. Summarization performed 
  by `MSstats` and `MSstatsTMT` packages is done separately for each ID in this column,
  - `PeptideSequence`, `PrecursorCharge`, `FragmentIon` and `ProductCharge`: these 
  four columns define a spectral feature (transition in SRM case). If information 
  for any of the columns is not available, it should be set to a constant value
  (for example `NA`),
  - `IsotopeLabelType`: column that indicates whether the measurement is based on an
  endogenous peptide (indicated by value "L" or "light") or reference peptide (indicated by value "H" or "heavy"),
  - `Run`: column that stores IDs of mass spectrometry runs. If annotation describing
  biological conditions and replicates is provided via a separate table, the run IDs
  should match `Run` IDs in the annotation,
  - `Condition`: column that stores labels for biological conditions (groups of interest).
  For time-course experiments, this column will represent time points. If the design
  experiment includes both time points and distinct biological subjects, these labels
  should be a combination of subject and time point,
  - `BioReplicate`: this column should contain a unique identifier for each biological replicate in the experiment. 
  For example, in a clinical proteomic investigation this should be a unique  patient ID. 
  Patients from distinct groups (indicated by the `Condition` column) should have distinct IDs,
  - `Intensity`: column that stores untransformed (in particular, no log transformation) measurements of feature abundance 
  in a given Run (and Channel in TMT case). 
  They can be peak heights, peak areas under the curve, or other quantitative representations
  of feature abundance,
  - for TMT datasets, a `Channel` column is required. Similarly to the `Run` column,
  values in this column must correspond to values in the annotation file, if provided separately.
  
Additionally, if the experiment involves fractionation, `Fraction` column can be added
to store fraction labels.

# Logging
  
`MSstatsConvert` allows for flexible logging based on the `log4r` package.
Information about preprocessing steps can be written to a file, to a console,
to both or to neither.
The `MSstatsLogsSettings` function helps manage log settings.
The user can pass a path to an existing file to the `log_file_path` parameter.
Combined with setting `append = TRUE`, this allows writing all information related to 
a specific data analysis to a single file.
If a user does not specify a file, a new file will be created automatically with a name
starting with "MSstats_log", followed by a timestamp.
  
```{r logs}
library(MSstatsConvert)
# default - creates a new file
MSstatsLogsSettings(use_log_file = TRUE, append = FALSE) 

# default - creates a new file
MSstatsLogsSettings(use_log_file = TRUE, append = TRUE, 
                    log_file_path = "log_file.log") 

# switches logging off
MSstatsLogsSettings(use_log_file = FALSE, append = FALSE) 

# switches off logs and messages
MSstatsLogsSettings(use_log_file = FALSE, verbose = FALSE) 
```
  
Additionally, session info generated by the `utils::sessionInfo()` function
can be saved to file with the `MSstatsSaveSessionInfo` function.

```{r }
MSstatsSaveSessionInfo()
```
  
By default, the output file name will start with "MSstats_session_info" and end with 
a current timestamp.
  
# Importing and cleaning data
  
MS data processing by `MSstatsConvert` starts with importing and cleaning data.
The `MSstatsImport` function produces a wrapper for possibly multiple files that
may describe a single dataset.
For example, `MaxQuant` output consists of two files, while `OpenMS` outputs just
a single file.
  
```{r }
maxquant_evidence = read.csv(system.file("tinytest/raw_data/MaxQuant/mq_ev.csv",
                                         package = "MSstatsConvert"))
maxquant_proteins = read.csv(system.file("tinytest/raw_data/MaxQuant/mq_pg.csv",
                                         package = "MSstatsConvert"))
maxquant_imported = MSstatsImport(list(evidence = maxquant_evidence,
                                       protein_groups = maxquant_proteins),
                                  type = "MSstats", tool = "MaxQuant")
is(maxquant_imported)

openms_input = read.csv(system.file(
  "tinytest/raw_data/OpenMSTMT/openmstmt_input.csv",
  package = "MSstatsConvert"
))
openms_imported = MSstatsImport(list(input = openms_input),
                                "MSstatsTMT", "OpenMS")
is(openms_imported)
```
  
The `getInputFile` method allows user to retrieve the files:

```{r }
getInputFile(maxquant_imported, "evidence")[1:5, 1:5]
```
  
As a next step of the analysis, input files are combined into a single `data.table`
with standardized column names by the `MSstatsClean` function.
It is a generic function with built-in support for outputs of tools listed in 
the "Purpose of the MSstatsConvert package" section.
The `type` parameter is equal to either `MSstats` or `MSstatsTMT` and indicates 
if the data comes from a labelled TMT experiment.

For some datasets, `MSstatsClean` may require additional parameters described in the respective help files.
For our example datasets, the following calls merge input files into a single table.

```{r }
maxquant_cleaned = MSstatsClean(maxquant_imported, protein_id_col = "Proteins")
head(maxquant_cleaned)

openms_cleaned = MSstatsClean(openms_imported)
head(openms_cleaned)
```

If a user wants to use `MSstatsConvert` package with data in a format that is not
currently supported, it is enough to first re-format the data into a `data.table`
with column ProteinName, PeptideSequence, PrecursorCharge, FragmentIon, ProductCharge (with the latter two possibly equal to NA), Run and IsotopeLabelType (in case of non-TMT data) or 
Channel (in case of TMT data). Moreover, the dataset may include any column that will be used for filtering the dataset (for example a column that stores q-values).
In our example, such additional columns are "Modifications" and "Score" from MaxQuant files.

Annotation columns should be called Condition and BioReplicate. For TMT data,
Mixture, TechRepMixture columns may be added.
Fractionation should be indicated by a Fraction column.

# Preprocessing
  
The goal of `MSstatsPreprocess` function is to transform cleaned MS data into a format
ready for statistical analysis with `MSstats` or `MSstatsTMT` packages.
This function accepts several parameters, and each corresponds to a preprocessing step.

  - `input` parameter is the dataset for preprocessing,
  - `annotation` is a description of biological conditions and replicates associated with
  MS runs (and channels in TMT case). If annotation is already included in the `input`,
  it should be equal to `NULL`. The annotation should be created by the
  `MSstatsMakeAnnotation` function,
  - `feature_columns` is a vector of column names that will denote features,
  - `remove_shared_peptides` is a logical parameter - if `TRUE`, shared peptides 
  will be removed from the analysis. Currently, `MSstats` assumes that only unique peptides
  are used, and presence of shared peptides may cause issues,
  - `remove_single_feature_proteins` is a logical parameter that indicates if proteins
  that only have a single feature should be removed from the analysis (`TRUE`),
  - `feature_cleaning` is a list, that currently consists of two named elements:
  `remove_features_with_few_measurements` should be equal to TRUE or FALSE. In the first case,
  feature that have less than three measurements across runs (or channels in a run for TMT data)
  will be removed. FALSE means that only features with no non-missing measurements will be removed. The `summarize_multiple_psms` element should be a function that will be used to aggregate multiple feature measurements within a single MS run,
  - `aggregate_isotopic` is a logical parameter - `TRUE` means that isotopic peaks will be aggregated (currently only used for Skyline input),
  - `columns_to_fill` is an optional named list with names corresponding to columns 
  and values correponding to values that will be used for these columns. For example,
  if the dataset is missing information about `ProductCharge`, such a column can be 
  added by passing `list(ProductCharge = NA)` to this parameter,
  - `score_filtering`, `exact_filtering` and `pattern_filtering` parameters are optional parameters that can be used to perform data filtering. An example is given below.
  

```{r }
maxquant_annotation = read.csv(system.file(
  "tinytest/raw_data/MaxQuant/annotation.csv",
  package = "MSstatsConvert"
))
maxquant_annotation = MSstatsMakeAnnotation(maxquant_cleaned,
                                            maxquant_annotation,
                                            Run = "Rawfile")
m_filter = list(col_name = "PeptideSequence", 
                pattern = "M", 
                filter = TRUE, 
                drop_column = FALSE)

oxidation_filter = list(col_name = "Modifications", 
                        pattern = "Oxidation", 
                        filter = TRUE, 
                        drop_column = TRUE)

feature_columns = c("PeptideSequence", "PrecursorCharge")
maxquant_processed = MSstatsPreprocess(
  maxquant_cleaned, 
  maxquant_annotation,
  feature_columns,
  remove_shared_peptides = TRUE, 
  remove_single_feature_proteins = FALSE,
  pattern_filtering = list(oxidation = oxidation_filter,
                           m = m_filter),
  feature_cleaning = list(remove_features_with_few_measurements = TRUE,
                          summarize_multiple_psms = max),
  columns_to_fill = list("FragmentIon" = NA,
                         "ProductCharge" = NA,
                         "IsotopeLabelType" = "L"))
head(maxquant_processed)

# OpenMS - TMT data
feature_columns_tmt = c("PeptideSequence", "PrecursorCharge")
openms_processed = MSstatsPreprocess(
  openms_cleaned, 
  NULL, 
  feature_columns_tmt,
  remove_shared_peptides = TRUE,
  remove_single_feature_proteins = TRUE,
  feature_cleaning = list(remove_features_with_few_measurements = TRUE,
                          summarize_multiple_psms = max)
)
head(openms_processed)
```
  
Annotation is created via the `MSstatsMakeAnnotation` function. It takes the cleaned dataset
and annotation file as input. Additionally, key-value pairs can be passed to this function
to change column names (not including dots and other symbols) in the annotation from names given by values to names given by keys.

For programmatic applications and consistency of the interface, filtering is done
with the help of lists.

For filtering based on numerical scores (for example q-value filtering), the list should 
consist of elements named 
  
  - `score_column`: name of a column that stores the score,
  - `score_threshold`: value above or below which measurements should be kept,
  - `direction`: if "greater", values greater than `score_threshold` will be kept; 
  if "smaller", values smaller than `score_threshold` will be kept;
  - `behavior`: if "remove", rows not below/above the threshold will be removed;
  if "replace", intensity in rows not below/above the threshold will be replaced by a given value,
  - `handle_na`: if "keep", `NA` in the score column will not be removed,
  - `fill_value`: value that will be used if `behavior = "replace"`,
  - `filter`: if `TRUE`, filtering will be performed (can be used for conditional filtering),
  - `drop_column`: if `TRUE`, column that stored the score will be removed.

For example, to remove intensities smaller than 1, we could pass the following list to
the `score_filtering` parameters:

```{r , eval = FALSE}
list(
  list(score_column = "Intensity", score_threshold = 1,
       direction = "greater", behavior = "remove", 
       handle_na = "remove", fill_value = NA, filter = TRUE, drop = FALSE
       )
)
```

For filtering based on patterns (for example, removing oxidation peptides), the list
should consist of elements named

  - `col_name`: name of a column that values that may be removed,
  - `filter_symbols`: vector of values - rows with these values in `col_name` will be removed
  or corresponding intensities will be replaced,
  - `behavior`: if "remove", rows that contain `filter_symbols` in `col_name` will be removed;
  if "replace", intensity in rows that contain `filter_symbols` in `col_name` will be replaced by a given value,
  - `fill_value`: value that will be used if `behavior = "replace"`,
  - `filter`: if `TRUE`, filtering will be performed (can be used for conditional filtering),
  - `drop_column`: if `TRUE`, column that stored the score will be removed.

For filtering based on exact values (for example, removing iRT proteins), the list
should consists of elements named

  - `col_name`: name of a column that stores strings that will be searched for given patterns,
  - `pattern`: vector of regular expressions - rows with matching values in `col_name` will be removed,
  - `filter`: if `TRUE`, filtering will be performed (can be used for conditional filtering),
  - `drop_column`: if `TRUE`, column that stored the values for filtering will be removed.

# Fractions and balanced design

Finally, after preprocessing, `MSstatsBalancedDesign` function can be applied to 
handle fractions and create balanced design.
For label-free and SRM data, it means that fractionation or technical replicates will be detected if these information is not provided. Features measured in multiple fractions (overlapped) will be assigned to a unique fraction. Then, the data will be adjusted so that within each fraction, every feature has a row for certain run. If the intensity value is missing, it will be denoted by `NA`.

For TMT data, a unique fraction will be selected for each overlapped feature and the 
data will adjusted so that within each run, every feature has a row for each channel.
If the intensity is missing for a channel, it will be denoted by `NA`.

```{r }
maxquant_balanced = MSstatsBalancedDesign(maxquant_processed, feature_columns)
head(maxquant_balanced)
dim(maxquant_balanced)
dim(maxquant_processed)

openms_balanced = MSstatsBalancedDesign(openms_processed, feature_columns_tmt)
head(openms_balanced)
dim(openms_balanced)
dim(openms_processed)
```

`MSstatsBalancedDesign` output is a `data.frame` of class `MSstatsValidated`.
Such a `data.frame` will be recognized by statistical processing functions from
`MSstats` and `MSstatsTMT` packages as a valid input, which will allow them to skip
checks and transformation necessary to fit data into this format.

